---
title: "Reproducible Example"
author: "Isaac Baguisa"
date: "2025-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

This is a minimal reproducible example to replicate some parts of my analysis, and build the Gaussian Process Regression and One-Class Support Vector Machine models to detect stellar flares.  

# Step 1: Read and Process Data

**Load required libraries**
```{r}
library(tidyverse)
library(kernlab)
library(e1071)
library(forecast)
library(ggplot2)
library(imputeTS)
library(extraDistr)
```

**Get data from GitHub**

```{r}
# Download data from GitHub as .zip file
repo_url <- "https://github.com/isaacbaguisa/Detecting-Stellar-Flares"
zip_url <- paste0(repo_url, "/archive/refs/heads/main.zip")
zip_file <- "stellar-flares.zip"
unzip_dir <- "stellar-flares"
download.file(zip_url, destfile = zip_file, mode = "wb")

# Unzip file
if (!dir.exists(unzip_dir)) {
  unzip(zip_file, exdir = ".")
  unzipped_name <- list.files(pattern = "^Detecting-Stellar-Flares-(main|master)$")
  file.rename(unzipped_name, unzip_dir)
}
```

**Read and impute data**
```{r}
# Read in
s1 <- read_csv("stellar-flares/Data/031381302.csv")
s2 <- read_csv("stellar-flares/Data/129646813.csv")
s3 <- read_csv("stellar-flares/Data/0131799991.csv")

# Imputation
s1$pdcsap_flux_imputed <- na_kalman(s1$pdcsap_flux, model = "StructTS")
s2$pdcsap_flux_imputed <- na_kalman(s2$pdcsap_flux, model = "StructTS")
s3$pdcsap_flux_imputed <- na_kalman(s3$pdcsap_flux, model = "StructTS")
```

# Step 2: Fit Gaussian Process Regression Model

```{r}
# Load GPR model directly because it takes a long time to run
load("stellar-flares/Data/gprModel.RData")
gprModel <- gpr_model

# Predict flux values using the GPR model
gprPred <- predict(gprModel, newdata = s1)

# Calculate residuals 
gprResiduals <- s1$pdcsap_flux_imputed - gprPred

# Define an anomaly threshold
threshold <- mean(gprResiduals, na.rm = TRUE) + 3 * sd(gprResiduals, na.rm = TRUE)

# Flag anomalies based on the residuals
s1$gpr_anomalies <- ifelse(gprResiduals > threshold, "Flare", "Normal")

# Create a data frame for plotting
plotData <- data.frame(time = s1$time, flux = s1$pdcsap_flux_imputed, flare = s1$gpr_anomalies)

# Visualize anomalies detected by GPR
ggplot(plotData, aes(x = time, y = flux)) +
  geom_point(aes(color = flare), size = 1) +
  scale_color_manual(values = c("Normal" = "black", "Flare" = "red")) +
  labs(title = "Gaussian Process Regression Anomalies",
       x = "Time", y = "PDCSAP Flux")

```

# Step 3: Fit One-Class Support Vector Machine

```{r}
# Fit the One-Class SVM model with decision function enabled
ocsvmModel <- svm(s1$pdcsap_flux_imputed, type = "one-classification", nu = 0.01,
                  kernel = "radial", scale = TRUE, 
                  gamma = 0.01, decision.values = TRUE)


# Get the decision function scores (distance from the boundary)
decisionScores <- ocsvmModel$decision.values

# Define a manual threshold for anomalies
threshold <- -quantile(decisionScores, 0.95) 

# Flag anomalies based on the threshold
s1$ocsvm_anomalies <- ifelse(decisionScores < threshold, "Flare", "Normal")

# Create a data frame for plotting
plotData <- data.frame(time = s1$time, flux = s1$pdcsap_flux_imputed, 
                       flare = s1$ocsvm_anomalies)
colnames(plotData) <- c("time", "flux", "flare")

# Visualize anomalies detected by OCSVM with manual threshold
ggplot(plotData, aes(x = time, y = flux)) +
  geom_point(aes(color = flare), size = 1) +
  scale_color_manual(values = c("Normal" = "black", "Flare" = "red")) +
  labs(title = "One-Class SVM Anomalies", 
       x = "Time", y = "PDCSAP Flux") 
```

# Step 4: Create SARIMA Simulation

Below is the function to simulate brightness
```{r}
# @param starData: A data frame containing the time series data of a star, 
# including flux values (with missing values imputed).
# @param starName: A string representing the name of the star (used in the plot title).
# @return A data frame containing the simulated light curve with injected 
# synthetic flares, including the time, simulated flux, and flare labels.

simulate_light_curve_pareto <- function(star_data, star_name) {
  
  # Fit the SARIMA model (or another model if needed)
  fit_sarima <- auto.arima(star_data$pdcsap_flux_imputed, seasonal = TRUE)
  
  # Simulate the light curve using the SARIMA model
  simulated_data <- simulate(fit_sarima, nsim = 
                               length(star_data$pdcsap_flux_imputed))
  
  # Initialize simulated dataframe
  simulated_df <- data.frame(
    time = 1:length(star_data$pdcsap_flux_imputed),
    pdcsap_flux_imputed = simulated_data,
    flare_label = rep(0, length(star_data$pdcsap_flux_imputed))  # Initially no flares
  )
  
  # Inject 10 synthetic flares using a Pareto distribution
  flare_indices <- sample(1:length(star_data$pdcsap_flux_imputed), 
                          size = 10, replace = FALSE)
  
  for (index in flare_indices) {
    # Use the Pareto distribution to determine flare amplitude
    flare_amplitude <- rpareto(1, a = 1, b = 25)  # Using parameters for large flares
    simulated_df$pdcsap_flux_imputed[index] <- simulated_df$pdcsap_flux_imputed[index] + 
      flare_amplitude
    simulated_df$flare_label[index] <- 1  # Mark flare
  }
  
  # Plot the actual data and simulated data
  p <- ggplot() +
    geom_line(data = star_data, aes(x = 1:length(star_data$pdcsap_flux_imputed), 
                                    y = star_data$pdcsap_flux_imputed), 
              color = "blue", alpha = 0.7) +  # Actual data
    geom_line(data = simulated_df, aes(x = time, y = pdcsap_flux_imputed), 
              color = "green", alpha = 0.7) +  # Simulated data
    geom_point(data = simulated_df[simulated_df$flare_label == 1, ], 
               aes(x = time, y = pdcsap_flux_imputed), color = "red") +  # Flares
    labs(title = paste("Actual vs. Simulated Light Curve for", star_name),
         x = "Time", y = "PDCSAP Flux") +
    theme_minimal() +
    scale_color_manual(values = c("blue", "green"))
  
  print(p)
  
  return(simulated_df)
}
```

I will use the same simulations as in my report.
```{r}
load("../Simulations/simulatedS1.RData")
load("../Simulations/simulatedS3.RData")
```

# Step 5: Score and Visualize GPR model on TIC 031381302 and TIC 0131799991

## Scoring GPR Trained on TIC 031381302 on TIC 031381302
```{r}
# Predict using the trained GPR model on the simulated data
gprPred <- predict(gprModel, newdata = data.frame(time = simulatedS1$time))

# Calculate residuals 
gprResiduals <- simulatedS1$pdcsap_flux_imputed - gprPred

# Define an anomaly threshold (residuals greater than 3 sd from the mean)
thresholdGpr <- mean(gprResiduals, na.rm = TRUE) + 3 * sd(gprResiduals, 
                                                          na.rm = TRUE)

# Flag anomalies based on the residuals
simulatedS1$predicted_flare_gpr <- ifelse(gprResiduals > thresholdGpr, 1, 0)  

# Calculate TN, TP, FP, FN for GPR
predictedFlaresGpr <- sum(simulatedS1$predicted_flare_gpr==1)
truePositivesGpr <- sum(simulatedS1$predicted_flare_gpr == 1 & 
                          simulatedS1$flare_label == 1)
falseNegativesGpr <- sum(simulatedS1$predicted_flare_gpr == 0 & 
                           simulatedS1$flare_label == 1)
falsePositivesGpr <- sum(simulatedS1$predicted_flare_gpr == 1 & 
                           simulatedS1$flare_label == 0)
trueNegativesGpr <- sum(simulatedS1$predicted_flare_gpr == 0 & 
                          simulatedS1$flare_label == 0)

# Calculate FPR, Sensitivity and Accuracy for GPR
falsePostiveRateGpr <- falsePositivesGpr / (falsePositivesGpr + trueNegativesGpr)
sensitivityGpr <- truePositivesGpr / (truePositivesGpr + falseNegativesGpr)
accuracyGpr <- mean(simulatedS1$predicted_flare_gpr == simulatedS1$flare_label)

# Print the results
cat("GPR Flares Predicted:", predictedFlaresGpr, "\n")
cat("GPR Sensitivity (True Positive Rate): ", sensitivityGpr, "\n")
cat("GPR Specificity: ", 1-falsePostiveRateGpr, "\n")
cat("GPR Predictive Accuracy: ", accuracyGpr, "\n")
```

## Scoring GPR Trained on TIC 031381302 on TIC 0131799991
```{r}
gprPred <- predict(gprModel, newdata = data.frame(time = simulatedS3$time))

# Calculate residuals 
gprResiduals <- simulatedS3$pdcsap_flux_imputed - gprPred

# Define an anomaly threshold (residuals greater than 3 sd from the mean)
thresholdGpr <- mean(gprResiduals, na.rm = TRUE) + 3 * sd(gprResiduals, 
                                                          na.rm = TRUE)

# Flag anomalies based on the residuals
simulatedS3$predicted_flare_gpr <- ifelse(gprResiduals > thresholdGpr, 1, 0)  

# Calculate TN, TP, FP, FN for GPR
predictedFlaresGpr <- sum(simulatedS3$predicted_flare_gpr==1)
truePositivesGpr <- sum(simulatedS3$predicted_flare_gpr == 1 & 
                          simulatedS3$flare_label == 1)
falseNegativesGpr <- sum(simulatedS3$predicted_flare_gpr == 0 & 
                           simulatedS3$flare_label == 1)
falsePositivesGpr <- sum(simulatedS3$predicted_flare_gpr == 1 & 
                           simulatedS3$flare_label == 0)
trueNegativesGpr <- sum(simulatedS3$predicted_flare_gpr == 0 & 
                         simulatedS3$flare_label == 0)

# Calculate FPR, Sensitivity and Accuracy for GPR
falsePostiveRateGpr <- falsePositivesGpr / (falsePositivesGpr + trueNegativesGpr)
sensitivityGpr <- truePositivesGpr / (truePositivesGpr + falseNegativesGpr)
accuracyGpr <- mean(simulatedS3$predicted_flare_gpr == simulatedS3$flare_label)

# Print the results
cat("GPR Flares Predicted:", predictedFlaresGpr, "\n")
cat("GPR Sensitivity (True Positive Rate): ", sensitivityGpr, "\n")
cat("GPR Specificity: ", 1-falsePostiveRateGpr, "\n")
cat("GPR Predictive Accuracy: ", accuracyGpr, "\n")

```

## Visualize GPR Flare Detection
```{r}
ggplot(simulatedS3, aes(x = time, y = pdcsap_flux_imputed)) + 
  geom_line(color = "green") +
  geom_point(data = subset(simulatedS3, predicted_flare_gpr == 1),
             aes(x = time, y = pdcsap_flux_imputed),
             color = "red", size = 2) +
  labs(title = "Flare Detection using GPR on TIC 0131799991 Simulation",
       x = "Time", y = "PDCSAP_FLUX") + 
         theme_minimal()

```

# Step 6: Score and visualize OCSVM model on TIC 031381302 and TIC 0131799991

## Scoring OCSVM Trained on TIC 031381302 on TIC 031381302
```{r}
ocsvmModel <- svm(s1$pdcsap_flux_imputed, type = "one-classification", nu = 0.01,
                  kernel = "radial", scale = TRUE, 
                  gamma = 0.01, decision.values = TRUE)

# Predict using the trained OCSVM model on the simulated data for Star 1
ocsvmPred <- predict(ocsvmModel, 
                     newdata = data.frame(simulatedS1$pdcsap_flux_imputed))

# Define threshold
simulatedS1$predicted_flare_ocsvm <- ifelse(ocsvmPred, 0, 1)
threshold <- quantile(simulatedS1$pdcsap_flux_imputed, 0.95)  
simulatedS1$predicted_flare_ocsvm <- ifelse(simulatedS1$predicted_flare_ocsvm == 1 & 
                                        simulatedS1$pdcsap_flux_imputed > threshold, 1, 0)

# Calculate TN, TP, FN, FP OCSVM
flaresPredictedOcsvm <- sum(simulatedS1$predicted_flare_ocsvm == 1)
truePositivesOcsvm <- sum(simulatedS1$predicted_flare_ocsvm == 1 & 
                            simulatedS1$flare_label == 1)
falseNegativesOcsvm <- sum(simulatedS1$predicted_flare_ocsvm == 0 & 
                             simulatedS1$flare_label == 1)
falsePositivesOcsvm <- sum(simulatedS1$predicted_flare_ocsvm == 1 & 
                             simulatedS1$flare_label == 0)
trueNegativesOcsvm <- sum(simulatedS1$predicted_flare_ocsvm == 0 & 
                          simulatedS1$flare_label == 0)

# Calculate FPR, Sensitivity and Accuracy for OCSVM
falsePostiveRateOcsvm <- falsePositivesOcsvm / (falsePositivesOcsvm + trueNegativesOcsvm)
sensitivityOcsvm <- truePositivesOcsvm / (truePositivesOcsvm + falseNegativesOcsvm)
accuracyOcsvm <- mean(simulatedS1$predicted_flare_ocsvm == simulatedS1$flare_label)

# Print the results
cat("OCSVM Flares Predicted:", flaresPredictedOcsvm, "\n")
cat("OCSVM Sensitivity (True Positive Rate): ", sensitivityOcsvm, "\n")
cat("OCSVM Specificity: ", 1-falsePostiveRateOcsvm, "\n")
cat("OCSVM Predictive Accuracy: ", accuracyOcsvm, "\n")
```

## Scoring OCSVM Trained on TIC 0131799991 on TIC 0131799991
```{r}
ocsvmModel <- svm(s3$pdcsap_flux_imputed, type = "one-classification", nu = 0.01,
                 kernel = "radial", scale = TRUE, 
                  gamma = 0.01, decision.values = TRUE)

# Predict using the trained OCSVM model on the simulated data for Star 3
ocsvmPred <- predict(ocsvmModel, 
                     newdata = data.frame(simulatedS3$pdcsap_flux_imputed))

simulatedS3$predicted_flare_ocsvm <- ifelse(ocsvmPred, 0, 1)

threshold <- quantile(simulatedS3$pdcsap_flux_imputed, 0.95)  

simulatedS3$predicted_flare_ocsvm <- ifelse(simulatedS3$predicted_flare_ocsvm == 1 & 
                                              simulatedS3$pdcsap_flux_imputed > threshold, 1, 0)

# Calculate TN, TP, FP, FN for OCSVM
flaresPredictedOcsvm <- sum(simulatedS3$predicted_flare_ocsvm == 1)
truePositivesOcsvm <- sum(simulatedS3$predicted_flare_ocsvm == 1 & 
                            simulatedS3$flare_label == 1)
falseNegativesOcsvm <- sum(simulatedS3$predicted_flare_ocsvm == 0 & 
                             simulatedS3$flare_label == 1)
falsePositivesOcsvm <- sum(simulatedS3$predicted_flare_ocsvm == 1 & 
                             simulatedS3$flare_label == 0)
trueNegativesOcsvm <- sum(simulatedS3$predicted_flare_ocsvm == 0 & 
                            simulatedS3$flare_label == 0)

# Calculate FPR, Sensitivity and Accuracy for GPR
falsePostiveRateOcsvm <- falsePositivesOcsvm / (falsePositivesOcsvm + trueNegativesOcsvm)
sensitivityOcsvm <- truePositivesOcsvm / (truePositivesOcsvm + falseNegativesOcsvm)
accuracyOcsvm <- mean(simulatedS3$predicted_flare_ocsvm == simulatedS3$flare_label)

# Print the results
cat("OCSVM Flares Predicted:", flaresPredictedOcsvm, "\n")
cat("OCSVM Sensitivity (True Positive Rate): ", sensitivityOcsvm, "\n")
cat("OCSVM Specificity: ", 1-falsePostiveRateOcsvm, "\n")
cat("OCSVM Predictive Accuracy: ", accuracyOcsvm, "\n")
```

## Visualize OCSVM Flare Detection
```{r}
# Visualize flux over time with anomalies highlighted
ggplot(simulatedS3, aes(x = time, y = pdcsap_flux_imputed)) + 
  geom_line(color = "green") +
  geom_point(data = subset(simulatedS3, predicted_flare_ocsvm == 1),
             aes(x = time, y = pdcsap_flux_imputed),
             color = "red", size = 2) +
  labs(title = "Flare Detection using One-Class SVM on TIC 0131799991 Simulation",
       x = "Time", y = "PDCSAP_FLUX") +
  theme_minimal()

```

